---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/JonahVentures/MitoPilot/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JonahVentures/MitoPilot/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

# Overview

MitoPilot is a package for the assembly and annotation of mitochondrial genomes from genome skimming data. The core application consists of a [Nextflow](https://www.nextflow.io/docs/latest/index.html) pipeline that is wrapped in an R package, which includes an R-Shiny graphical interface to monitor and interact with processing parameters and outputs. Currently the pipeline expects paired-end Illumina reads as the raw input and performs the following steps:

1. Mitogenome assembly
   - [fastp](https://github.com/OpenGene/fastp) for quality control and adapter trimming
   - [GetOrganelle](https://github.com/Kinggerm/GetOrganelle) for mitogenome assembly
   - [bowtie2](https://github.com/BenLangmead/bowtie2) for read mapping to calculate coverage and error rates.
2. Mitogenome annotation 
   - [MITOS2](https://gitlab.com/Bernt/MITOS) for rRNA and PCG annotation
   - [tRNAscan-SE](https://github.com/UCSC-LoweLab/tRNAscan-SE) for tRNA annotation
   - Custom scripts for gene boundary refinement and annotation file formatting
   - Validation to flag possible issues or known errors that would be rejected by NCBI GenBank
   - Manual curation of annotations using the integrated Shiny App.
3. Data export
   - Custom scripts to export data in a format suitable for submission to NCBI GenBank
   
MitoPilot is currently optimized for Fish mitogenome assembly, but has been developed with modularity and extensibility in mind to facilitate broader application in the future. In particular, the included reference databases for assembly and annotation, along with the default parameters for annotation curation and validation are highly specific to fish mitogenomes. However, the reference databases and annotation parameters can be easily updated. However, the custom logic in the annotation curation and validation scripts would likely need to be updated for optimal performance with other taxonomic groups. Because all of the dependencies and reference data is contained on the underlying Docker Image (currently hosted at `drleopold/MitoPilot`), customization or extension will involve updating the Docker image appropriately and specifying the new image in the Nextflow configuration file (see below). The Dockerfile and default reference database file are included in this repository for reference and a custom local Docker Image can be generated by modifying the Dockerfile as needed and running `./docker/deploy-local.sh latest` in the repository root directory. More detailed instructions on using custom reference databases (along with other advanced customization) are coming soon.

# Installation

To use MitoPilot, you will need [R (>=4.4.0)](https://www.r-project.org/) and [Nextflow](https://www.nextflow.io/docs/latest/install.html). In addition, depending or where Nextflow will be executing the pipeline (e.g., locally or on a remote cluster), you may also need to install [Docker](https://docs.docker.com/engine/install/) or [Singularity](https://docs.sylabs.io/guides/latest/user-guide/quick_start.html#quick-installation-steps).

Once you have R and Nextflow installed, install `{MitoPilot}` in R from GitHub:

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("JonahVentures/MitoPilot")
```

Alternatively, you can clone this repository and install the package locally from the project folder:

```{r, eval = FALSE}
devtools::install()
```

# Usage

MitoPilot includes a set of pre-filtered test data and a function for setting up an example project, `new_test_project()`. It is highly recommended that you use the test project to ensure successful installation and familiarize yourself with the pipeline before running on your own data. A complete tutorial on running the test project is available in the package vignette, [here](https://jonahventures.github.io/MitoPilot/articles/test-project.html).

## Initializing A Project

The MitoPilot workflow begins by initializing a new project with the `new_project()` function. If running from within RStudio (recommended) a new R-project will also be initialized and opened in a new RStudio session. 

```{r, eval = FALSE}
MitoPilot::new_project(
  path = "path/to/project",
  mapping_fn = "path/to/mapping_file.csv",
  data_dir = "path/to/raw_data",
  executor = "local"
)
```

+ Path 
   - The path specifies where the new project directory will be created. If no path is provided, the project will be created in the current working directory. 
+ Mapping File
   - The mapping file should be in csv format and must contain the columns ID (a unique identifier for each sample), Taxon (eg, species or genus name), along with R1, and R2 (specifying the forward and revers file names for the raw Illumina paired end data). In addition to the required fields any other sample metadata can be included in the mapping file. These fields can also be used when exporting files for NCBI GenBank Submissions, so metadata that is important for submission (e.g., BioSample ID) can be included here. 
+ Data Directory
   - The data directory should contain the raw Illumina paired-end reads specified in the mapping file. 
+ Executor
   - The executor specifies where the computational work will be performed by Nextflow. For example choosing `local` will run the pipeline on the local machine, while `awsbatch` will run the pipeline on AWS Batch. Running `new_project()` will generate a executor-specific .config file in the project directory that must be edited to specify additional parameters for the pipeline to run.

### Nextflow Configuration File

Initializing a new project will populate the `.config` file in the project directory that may include place holders for important parameters, in the format: `<<PARAMETER_NAME>>`. For example, all new configuration files will include the line `rawDir = '<<RAW_DIR>>'`, which should be  updated to `rawDir = '/path/to/your/data'` indicating the location of the raw data file specified in the mapping file. The configuration files can also be modified to specify custom docker images for one or more of the processing steps. After initializing a new project you should review the `.config` file to ensure that all necessary parameters are provided. 

### Database Creation

MitoPilot makes use of the Nextflow plugin [nf-sqldb](https://github.com/nextflow-io/nf-sqldb) to store and retrieve processing parameters and information about the samples and their processing status. The database (`.sqlite`) is created automatically when the project is initialized and is stored in the project directory. The interactive MitoPilot GUI also interacts with this database to allow you run the pipeline, modify parameters, and view the results. When initializing a new project, default processing parameters for the pipeline modules are stored in the database, but any processing parameters can also be passed to the `new_project()` function to modify the initial defaults. For example, the following options would modify the allocated memory and GetOrganelle command line options :

```{r, eval = FALSE}
MitoPilot::new_project(
  mapping = "path/to/mapping_file.csv",
  executor = "local",
  assemble_memory = 24,
  getOrganelle = "-F 'anonym' -R 20 -k '21,45,65,85,105,115' -J 1 -M 1 --expected-max-size 20000 --target-genome-size 16500"
)
```

For complete list of available parameters that can be set during project initialization, see the `new_db()` function documentation. 

Although the MitoPilot GUI provides an interface to the database, during troubleshooting it is often helpful to directly explore the contents of the project's `.sqlite` database. This can be easily done in R using the `{dplyr}` extension, [{dbplyr}](https://dbplyr.tidyverse.org/), which is used extensively in the MitoPilot package, along with [{DBI}](https://dbi.r-dbi.org/), for database interactions. Alternatively, many interactive tools exist specifically for working with SQLite databases, such as [DB Browser for SQLite](https://sqlitebrowser.org/). 

### Database Modification

MitoPilot databases can be modified using the R helper functions `update_sample_metadata()`, `update_sample_seqdata()`, and `add_samples()`. You must close any existing connections (e.g. the MitoPilot GUI) prior to modifying the database. These functions will automatically create backups of the database in case you need to revert your changes. For more information, please see the [manual pages](https://jonahventures.github.io/MitoPilot/reference/index.html) for these functions.

# Running The Pipeline

Once a project is initialized, the pipeline status caa be viewed using the MitoPilot GUI. The GUI can be launched by running the `MitoPilot()` command in the R console from the project directory. The GUI will open in a new browser window and is primarily comprised of an interactive table, with 3 modules (Assembly, Annotate, Export), where each row represents a sample in the project. 

Please note that we have tested the MitoPilot GUI on Chrome and Firefox. There are some known bugs when running the GUI on Safari, it may be best to avoid this browser for MitoPilot.

## Sample Status

In the Assemble and Annotate modules the icon at the start of each row indicates the sample status, where:

1. (⏳) Hold / Waiting = Indicates that the sample is ready to be updated, but will not be updated the next time the pipeline is run.
2. (🏃) Ready to Run = Indicates that the sample will be updated the next time the pipeline is run.
3. (✅) Completed Successfully = Indicates that the sample has been successfully processed.
4. (⚠️) Completed with Warning - Processing is complete but may have failed or needs manual review.

There is an additional icon indicating whether a samples is locked () or unlocked (). A locked sample will be protected from further updates by Nextflow. Locking a sample will also make it available in the next MitoPilot module - a sample must be locked in the Assemble module to proceed with Annotation and must be locked the the Annotation modules to proceed with data Export. Both the "state" and "locked" status of one or more samples can be modified by selecting the sample rows in the table and using the "STATE" and "LOCK" buttons at the top of the interface. 

## Processing parameters

In the Assemble and Annotate modules, the processing parameters for one or more samples can be modified by clicking on the link in the relevant column (e.g., `Assemble Opts.`). This will open a popup that can be used to modify options by either selecting an existing option set from the drop-down menu, or by entering a new name for the option set and modifying the parameters. If multiple rows are selected in the table when the options popup is triggered, the changes will apply to all selected samples (though selecting any locked sample will prevent this action). An existing options set can also be modified by checking the "editing" box in the popup, but this may trigger a warning that the edits will affect more samples than are currently selected (i.e., all sample that are using that options set).

## Running Nextflow

When one or more samples are in the "Ready to Run" state, the Nextflow pipeline can be run by clicking the "UPDATE" button at the top of the interface. This will open a popup where the `Start Nextflow` button can be pressed and output from the pipeline can be viewed to track progress. Alternatively, the Nextflow command displayed in the popup can be copied and run in the a terminal from the project directory, which can be useful if you would like to specify additional command line options or override input parameters. 

# Development Notes

  - This package uses [{renv}]() for package management. After cloning the repository, run `renv::restore()` to install the necessary packages.
  - To work from the package repository, but reference a MitoPilot project in a different directory, set the `MitoPilot.db` option to the location of the `.sqlite` database for the project (e.g. `options("MitoPilot.db" = "~/Jonah/MitoPilot-testing/.sqlite")`).
  - When modifying the underlying R-package functions references in the Nextflow pipeline, or modifying / adding reference databases specified in `docker/Dockerfile`, the docker image should be rebuilt. The `docker/deploy-local.sh` script can be used to build a local image, or the `docker/deploy-aws.sh` and `docker/deploy-dockerhub.sh` scripts can be modified to deploy a remote image to your account. In any case, the Nextflow `.config` file should be modified such that one or more of the processing steps reference the new image.





