---
title: "Custom HPC Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom HPC Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Running MitoPilot on your own HPC cluster

MitoPilot ships built-in executor templates for `local`, `awsbatch`, the
Smithsonian Hydra cluster (`NMNH_Hydra`), and the NOAA SEDNA cluster
(`NOAA_SEDNA`). For any other cluster, MitoPilot includes **generic templates**
for the four most common schedulers, plus a helper function to build a Nextflow 
config for your cluster once and reuse it for every project.

| Scheduler | `scheduler =` | Nextflow executor |
|-----------|---------------|-------------------|
| SLURM | `"slurm"` | `slurm` |
| SGE / UGE | `"sge"` | `sge` |
| PBS Pro / OpenPBS / Torque | `"pbs"` | `pbspro` (edit to `pbs` for Torque/OpenPBS) |
| IBM Spectrum LSF | `"lsf"` | `lsf` |

## Configure MitoPilot for your cluster

`generate_config()` builds a Nextflow config from a generic template, fills in
your cluster-specific settings, and saves it as a **named profile** in your
persistent MitoPilot config directory (see `mitopilot_config_dir()`).

```{r eval=FALSE}
library(MitoPilot)

generate_config(
  name             = "my_cluster",     # what you'll pass to new_project(executor = ...)
  scheduler        = "slurm",
  queue            = "general",         # partition / queue (omit to use the cluster default)
  account          = "my_allocation",   # folded into clusterOptions (SLURM --account, PBS -A, LSF -P, SGE -P)
  cluster_options  = "--qos=long",      # any extra raw scheduler directives (optional)
  container_engine = "apptainer",       # "singularity", "apptainer", or "docker"
  container_cache  = "/scratch/$USER/sif" # optional cacheDir for singularity/apptainer
)
```

This writes `config.my_cluster` to your config directory. Prefer to be prompted
interactively? Pass `interactive = TRUE` and leave the cluster values unset.

## Reuse your cluster config for every project

Once saved, the profile name behaves like any built-in executor:

```{r eval=FALSE}
new_project(
  path       = "path/to/project",
  mapping_fn = "path/to/mapping.csv",
  data_path  = "path/to/raw_data",
  executor   = "my_cluster"
)
```

`new_project()` resolves the saved profile and fills in the per-project values
(raw data directory, container image, genetic code, etc.), leaving a ready-to-run
`.config` in the project directory.

List everything available (built-in templates plus your saved profiles):

```{r eval=FALSE}
list_configs()
```

## Per-process resources

The generated config sets defaults for CPUs and memory that work for most
datasets. Per-step resource requests (`coverage`, `blast_ref_align`,
`blast_gb`) live in the `params { }` block of the `.config` file and can be
edited by hand. Memory and CPU for all other steps in the workflow are pulled dynamically
from the project database and can be set directly in the MitoPilot app.

## Validate before running

If `nextflow` is on your `PATH`, you can sanity-check the generated config:

```{bash eval=FALSE}
cd path/to/project
cp .config nextflow.config
nextflow config .          # prints the resolved config, or a parse error
rm nextflow.config
```

After reviewing the `.config` file, launch the pipeline from the MitoPilot app
or submit it as a batch job using your cluster's normal submission tools (e.g.
`sbatch`, `qsub`, `bsub`) wrapping the `nextflow run` command. See the
[NMNH Hydra](NMNH-Hydra.html) and [NOAA SEDNA](NOAA-SEDNA.html) vignettes for
batch submission examples you can adapt.

## Run everything from the MitoPilot container (recommended)

You do not need to install R, or any of the bioinformatics tools, on the
cluster. The MitoPilot Docker image is a single self-contained "box" that
already includes R, MitoPilot, and every tool the pipeline uses. You can use the
R *inside* that box to create projects, configure your cluster, and run the
headless GUI.

A quick vocabulary primer if containers are new to you:

- **Docker image**: the prebuilt box. MitoPilot publishes one on Docker Hub.
- **Singularity / Apptainer**: the container runtime that nearly every HPC
  cluster provides (Docker itself is usually not permitted on shared clusters).
  It can run a Docker image directly. The two names are interchangeable here.
- **`.sif` file**: Singularity's single-file copy of the image. You build it
  once, then reuse it.

### 1. Pull the image once

From an interactive session or compute node (not the login node):

```{bash eval=FALSE}
# match the tag to your installed MitoPilot version
singularity pull mitopilot.sif docker://macguigand/mitopilot:1.4.6
```

You may see many lines like `... ignoring (usually) harmless EPERM on setxattr
...`. These are harmless rootless-conversion warnings; the `.sif` still builds.

### 2. Understand directory binding

By default a Singularity container can only see your **home directory**.
Anything under home is available automatically. If your project, raw data, or
scratch space lives somewhere else (for example `/scratch` or `/pool`), you must
**bind** those paths with `--bind` so the container can read and write them:

```{bash eval=FALSE}
singularity exec --bind /scratch,/pool mitopilot.sif R
```

> **Saved cluster configs need no extra binding.** `generate_config()` writes
> its profiles to `~/.config/R/MitoPilot` inside your home directory, and
> Singularity mounts your home automatically. A profile you save once therefore
> persists and is found by `new_project()` on every later run, with no `--bind`
> needed. (This only breaks if you deliberately launch with `--contain` or
> `--no-home`, which detach the home directory.)

### 3. Create and configure a project with the bundled R

Use **absolute paths** and pin the working directory with `--pwd`. Singularity
sets the in-container working directory to your home by default, so a relative
path would create the project in the wrong place.

```{bash eval=FALSE}
cd /scratch/$USER/my_runs
singularity exec --bind /scratch --pwd $PWD mitopilot.sif R
```

```{r eval=FALSE}
library(MitoPilot)

# one-time: save a profile for your cluster (skip if using a built-in executor)
generate_config(name = "my_cluster", scheduler = "slurm", queue = "general")

new_project(
  path       = "/scratch/me/my_runs/project01",   # absolute path
  mapping_fn = "/scratch/me/my_runs/mapping.csv",
  data_path  = "/scratch/me/my_runs/raw_data",
  executor   = "my_cluster",                       # or a built-in, e.g. "NMNH_Hydra"
  container  = "macguigand/mitopilot:1.4.6"        # pipeline steps use the same image
)
```

### 4. Launch the headless GUI from the container

```{bash eval=FALSE}
proj=/scratch/$USER/my_runs/project01
singularity exec --bind /scratch --pwd "$proj" mitopilot.sif \
  R -e "setwd('$proj'); MitoPilot::MitoPilot(host='0.0.0.0', port=7591, launch.browser=FALSE)"
```

Then follow the SSH tunnel steps below to reach the GUI from your laptop.

## Headless GUI over an SSH tunnel

The MitoPilot GUI is just a Shiny web server over the project's SQLite database
and Nextflow. It needs **no browser and no X11 on the cluster**, so you do not
need RStudio Server just to use the GUI. Instead, run the server headless on a
cluster node with a fixed host and port, then reach it from your laptop browser
over an SSH tunnel.

**1. On the cluster**, from inside your project directory, launch MitoPilot
bound to all interfaces on a port of your choice and tell it not to open a
browser:

```{r eval=FALSE}
library(MitoPilot)
MitoPilot(host = "0.0.0.0", port = 7591, launch.browser = FALSE)
```

As the server starts it prints the exact tunnel command using the live node
hostname, for example:

```
ssh -N -L 7591:NODE:7591 <user>@<cluster>
```

**2. On your laptop**, open the tunnel (substitute your username and login
host; if MitoPilot is running on a compute node reached via `srun --pty`, the
printed `NODE` is that compute node and you may need an SSH jump through the
login node, `ssh -N -J <user>@<cluster> -L 7591:NODE:7591 <user>@NODE`):

```{bash eval=FALSE}
ssh -N -L 7591:NODE:7591 <user>@<cluster>
```

**3. In your browser**, open:

```
http://localhost:7591
```

Leave both the cluster R session and the local `ssh` session open while you use
the GUI. This replaces the RStudio Server requirement for users who only need
the GUI.

**Caveat:** in-app features that open a file or browser on the server side
(the "Open output folder" button, the HTML annotation report, and the ruleset
browser) will not work headless, since they try to open on the cluster rather
than your laptop. The main assemble / annotate / curate GUI is unaffected.

## Submitting the pipeline in headless mode

When MitoPilot runs headless it will **not** run Nextflow on the node hosting
the GUI, since that would tie up an interactive node or the login node. The
"Run from App" button is therefore disabled. Instead, when you start a workflow,
the update modal shows a ready-to-edit **cluster submission script**:

- It is pre-filled with the correct scheduler directives (`#SBATCH`, `#$`,
  `#PBS`, or `#BSUB`) for your project's executor.
- Edit the resources, and add your environment setup where indicated (for
  example `module load java`, `mamba activate ...`, or
  `export NXF_SINGULARITY_CACHEDIR=...`). Those lines are pre-filled as
  commented examples.
- Click **"Submit to Cluster"** to submit it with your scheduler's command
  (`sbatch` / `qsub` / `bsub`), or **"Save Script Only"** to write the script
  and submit it yourself.
- Your edits to the resource block are remembered per project (saved to
  `.mitopilot_submit.template`), so the next run pre-fills them. The "Resume
  previous run?" checkbox still adds or removes `-resume` without disturbing
  your edits.

> **Submitting from inside the container.** "Submit to Cluster" calls
> `sbatch` / `qsub` / `bsub` directly, and those commands are usually **not**
> available inside a Singularity container. If you launched the GUI from the
> container (the recommended setup above), use **"Save Script Only"** and then
> submit the written `.sh` from a normal cluster shell. The submitted job runs
> on a scheduler node (not inside the container) and calls Nextflow, which in
> turn uses Singularity for each pipeline step, so make sure `java` and
> `nextflow` are available via the script's environment-setup lines.
