---
title: "Custom HPC Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom HPC Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Running MitoPilot on your own HPC cluster

MitoPilot ships built-in executor templates for `local`, `awsbatch`, the
Smithsonian Hydra cluster (`NMNH_Hydra`), and the NOAA SEDNA cluster
(`NOAA_SEDNA`). For any other cluster, MitoPilot includes **generic templates**
for the four most common schedulers, plus a helper to configure your cluster
once and reuse it for every project.

| Scheduler | `scheduler =` | Nextflow executor |
|-----------|---------------|-------------------|
| SLURM | `"slurm"` | `slurm` |
| SGE / UGE | `"sge"` | `sge` |
| PBS Pro / OpenPBS / Torque | `"pbs"` | `pbspro` (edit to `pbs` for Torque/OpenPBS) |
| IBM Spectrum LSF | `"lsf"` | `lsf` |

## Configure your cluster once

`generate_config()` builds a Nextflow config from a generic template, fills in
your cluster-specific settings, and saves it as a **named profile** in your
persistent MitoPilot config directory (see `mitopilot_config_dir()`).

```{r eval=FALSE}
library(MitoPilot)

generate_config(
  name             = "my_cluster",     # what you'll pass to new_project(executor = ...)
  scheduler        = "slurm",
  queue            = "general",         # partition / queue (omit to use the cluster default)
  account          = "my_allocation",   # folded into clusterOptions (SLURM --account, PBS -A, LSF -P, SGE -P)
  cluster_options  = "--qos=long",      # any extra raw scheduler directives (optional)
  container_engine = "apptainer",       # "singularity", "apptainer", or "docker"
  container_cache  = "/scratch/$USER/sif" # optional cacheDir for singularity/apptainer
)
```

This writes `config.my_cluster` to your config directory. Prefer to be prompted
interactively? Pass `interactive = TRUE` and leave the cluster values unset.

## Reuse it for every project

Once saved, the profile name behaves like any built-in executor:

```{r eval=FALSE}
new_project(
  path       = "path/to/project",
  mapping_fn = "path/to/mapping.csv",
  data_path  = "path/to/raw_data",
  executor   = "my_cluster"
)
```

`new_project()` resolves the saved profile and fills in the per-project values
(raw data directory, container image, genetic code, etc.), leaving a ready-to-run
`.config` in the project directory.

List everything available (built-in templates plus your saved profiles):

```{r eval=FALSE}
list_configs()
```

## Per-process resources

The generated config sets defaults for CPUs and memory that work for most
datasets. Per-step resource requests (`coverage`, `blast_ref_align`,
`blast_gb`) live in the `params { }` block of the `.config` file and can be
edited by hand. Memory and CPU for the assembly steps are pulled dynamically
from the project database.

## Validate before running

If `nextflow` is on your `PATH`, you can sanity-check the generated config:

```{bash eval=FALSE}
cd path/to/project
cp .config nextflow.config
nextflow config .          # prints the resolved config, or a parse error
rm nextflow.config
```

After reviewing the `.config` file, launch the pipeline from the MitoPilot app
or submit it as a batch job using your cluster's normal submission tools (e.g.
`sbatch`, `qsub`, `bsub`) wrapping the `nextflow run` command. See the
[NMNH Hydra](NMNH-Hydra.html) and [NOAA SEDNA](NOAA-SEDNA.html) vignettes for
worked batch-submission examples you can adapt.
