---
title: "Installing Without Containers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing Without Containers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
.alert {
  border-left: 5px solid;
  padding: 10px;
  margin: 10px 0;
  border-radius: 5px;
}
.alert-tip { border-color: #28A745; background-color: #E9F7EF; }
.alert-note { border-color: #007BFF; background-color: #EBF5FF; }
.alert-warning { border-color: #FFC107; background-color: #FFF9E6; }
.alert-danger { border-color: #DC3545; background-color: #F8D7DA; }
/* Smithsonian palette: blue #009CDE, yellow #FFCD00, black #101820 */
.alert-hydra {
  border-color: #101820;
  background-color: #E7F6FC;
  color: #101820;
}
.alert-hydra .hydra-tag {
  display: inline-block;
  background-color: #FFCD00;
  color: #101820;
  font-size: 0.75em;
  font-weight: bold;
  letter-spacing: 0.04em;
  padding: 1px 7px;
  border-radius: 10px;
  margin-right: 6px;
  vertical-align: 2px;
}
strong { font-weight: bold; }
img { max-width: 100%; border: 1px solid #ddd; border-radius: 4px; }
.figure { margin: 1.2em 0; }
.figure .caption, p.caption {
  font-size: 0.82em;
  line-height: 1.35;
  color: #6c757d;
  font-style: italic;
  margin-top: 0.4em;
}
.part-break {
  border: none;
  border-top: 3px solid #c9d1d6;
  margin: 2.2em 0 1.6em;
}
</style>

# Running MitoPilot without Docker or Singularity

Some clusters do not allow Docker, Singularity, or Apptainer. MitoPilot can run
without them: every tool the pipeline needs installs into a folder you own with
a conda-style package manager, and no root access is required. This page sets
that up once per cluster.

<div class="alert alert-note">
<strong>What you get.</strong> The same pinned tool versions as the MitoPilot
container (R 4.5.2, Nextflow, GetOrganelle, MITOS2, BLAST+, and the rest), a
local copy of the metazoan BLAST database, and the MitoPilot R package, all under
one directory. Nothing uses user namespaces, ptrace, FUSE, or setuid binaries.
</div>

## Requirements

| Requirement | Notes |
|---|---|
| Linux x86_64, glibc 2.17 or newer | RHEL/Rocky/Alma 7+, Ubuntu 18.04+, Debian 10+ |
| About 20 GB of disk on a filesystem that allows executables | Home directories are often small or `noexec`; a project or scratch space is usually better |
| Outbound HTTPS from the node where you run the install script | conda-forge, bioconda, GitHub, and NCBI are contacted once |
| `curl`, `tar`, `bash` | Present on every cluster |
| Optional: an existing `conda`, `mamba`, or `pixi` | If absent, the install script downloads `micromamba` (a single binary) for you |

## 1. Run the install script

Download the install script and the environment definitions from the MitoPilot
release you want (the script and the R package must be the same version):

```{bash eval=FALSE}
ver=1.5.5
curl -L https://github.com/Smithsonian/MitoPilot/archive/refs/tags/${ver}.tar.gz | tar -xz
cd MitoPilot-${ver}/inst/native
```

Choose an install location on a large filesystem that allows executables. Where
that is depends on the cluster: a project, group, work, or scratch directory is
usually better than your home directory, and your cluster's documentation will
say where large software belongs. The examples on this page write `<install_dir>`
for that path. Pass it as `--prefix`; everything MitoPilot needs goes inside that
one directory:

```{bash eval=FALSE}
bash install_mitopilot_native.sh --prefix <install_dir>
```

That builds the core environments (30 to 90 minutes, mostly downloads), fetches
the BLAST database, and installs the MitoPilot R package into the bundled R.
Options:

| Flag | Effect |
|---|---|
| `--manager pixi` | Use pixi instead of micromamba; `mamba` and `conda` also accepted |
| `--no-optional` | Skip MitoFinder, ARWEN, and NCBI ORFfinder (saves about 2 GB; these are off by default in projects anyway) |
| `--skip-blast-db` | Skip the 289 MB database; reference search falls back to NCBI over the network |
| `--mitopilot-ref 1.5.5` | Install a MitoPilot tag from GitHub instead of the source tree the script came from |
| `--dry-run` | Show what would happen |

Re-running the script is safe; finished steps are skipped.

The install script sets strict channel priority (conda-forge over bioconda). If you
build the environments by hand from `inst/native/envs/*.yml`, pass
`--channel-priority strict` or set `CONDA_CHANNEL_PRIORITY=strict`.

**What the install directory contains.** `envs/` holds one environment per tool group
(`mitopilot` for R, Java, Nextflow, and the assembly tools, plus `mitos`,
`trnascan`, `aragorn`, `bamreadcount`, `orffinder`, and `mitofinder`), `ref_dbs/` the BLAST database, `opt/` the hand-built
optional tools, `nextflow_home/` Nextflow's plugin and engine cache, and
`activate.sh`. The environments are created by path, not by name, so
`conda env list` shows them without names. That is deliberate: everything
stays under one directory you can move or delete, and nothing lands among
your own named environments. You never activate them yourself;
`activate.sh` does.

This is a one-time install. Nothing in the install directory is rebuilt at run time,
and only a release that changes tool pins needs a re-run (see Updating).

## 2. Check the environment

From any R that has MitoPilot installed (the bundled one is at
`<install_dir>/envs/mitopilot/bin/R`):

```{r eval=FALSE}
library(MitoPilot)
native_check("<install_dir>")
```

Every core tool should show `found TRUE`. Optional tools are missing only if you
installed with `--no-optional`.

## 3. Save a config profile (once)

This step is needed for both a single workstation and a cluster. Without it,
`new_project(executor = "local")` uses the built-in template, which runs the
pipeline in Docker. The profile tells MitoPilot to use the install directory
instead, and it is saved so every later project can reuse it by name.

**Workstation or a single node, no scheduler:**

```{r eval=FALSE}
library(MitoPilot)
generate_config(
  name             = "local_native",
  scheduler        = "local",
  container_engine = "none",
  native_prefix    = "/path/to/mitopilot_native"   # the install location
)
```

**Cluster with a scheduler:**

```{r eval=FALSE}
generate_config(
  name             = "my_cluster",
  scheduler        = "slurm",             # or sge, pbs, lsf
  queue            = "general",
  account          = "my_allocation",
  container_engine = "none",
  native_prefix    = "<install_dir>" # the install location
)
```

**NMNH Hydra or NOAA SEDNA:** pass the built-in cluster template as
`scheduler` to keep its tuned resource settings (parallel environment,
memory flags, himem selection). This form is only for
`container_engine = "none"`; with containers, use
`new_project(executor = "NMNH_Hydra")` as before.

```{r eval=FALSE}
generate_config(
  name             = "hydra_native",
  scheduler        = "NMNH_Hydra",         # or NOAA_SEDNA
  container_engine = "none",
  native_prefix    = "<install_dir>"
)
```

Then create projects with `new_project(..., executor = "local_native")` or
`executor = "my_cluster"`, exactly as in
[HPC cluster support](Custom-HPC.html). The profile makes every pipeline task
source `<install_dir>/activate.sh` before it runs, so no module loads or conda
activation are needed on compute nodes.

## 4. Launch the app

Which R runs the **app** does not matter, as long as MitoPilot is installed in it
and it is version 4.4 or newer. Pipeline tasks always use the bundled R.

**From a terminal (SSH tunnel).** Source the environment, start R, and follow
[Accessing the MitoPilot app over an SSH tunnel](Custom-HPC.html#accessing-the-mitopilot-app-over-an-ssh-tunnel):

```{bash eval=FALSE}
source <install_dir>/activate.sh
cd /path/to/my_project
R -e "MitoPilot::MitoPilot(host='0.0.0.0', port=7591, launch.browser=FALSE)"
```

**From RStudio Server or Open OnDemand.** Install MitoPilot in that R once
(`BiocManager::install("Smithsonian/MitoPilot")`), then in each session:

```{r eval=FALSE}
library(MitoPilot)
native_setup("<install_dir>")
```

`native_setup()` puts the bundled Nextflow and Java on this session's PATH,
together with the job scheduler commands (`sbatch`, `qsub`, or `bsub`) and
their environment variables as they were when the install script ran, since
RStudio Server sessions do not inherit a login shell's PATH. If the scheduler
lives somewhere else on your cluster, edit the "job scheduler" lines near the
end of `activate.sh`. To
make it automatic, add `MITOPILOT_NATIVE_PREFIX=<install_dir>`
to `~/.Renviron` and call `native_setup()` with no arguments, or add the call to
`~/.Rprofile`. The **Start** button in the app reads the project config and finds
the environment on its own, so it works even without `native_setup()`; the setup
is what makes `library(MitoPilot)` report the right Nextflow version and lets you
run `nextflow` from the R console.

**Batch submission.** When the app offers a cluster submission script, the line
`source <install_dir>/activate.sh` is already filled in.

**On a workstation.** With `activate.sh` sourced, `MitoPilot()` from the bundled
R opens your browser through `xdg-open`, which the bundled R uses because it
ships with no browser setting of its own. On a headless machine there is
nothing to open; pass `launch.browser = FALSE` and use the tunnel above.

<div class="alert alert-warning">
<strong>Warning.</strong> A user-level <code>NXF_VER</code> (for example in
<code>~/.Renviron</code> or <code>~/.bashrc</code>) overrides the bundled
Nextflow in every R session. The <strong>Start</strong> button and the submit
script pin the bundled version themselves, but <code>library(MitoPilot)</code>
will report the other version and console calls to <code>nextflow</code> will
download it. Remove the setting on a cluster that uses the native install.
</div>

## Updating

Update the R package inside the bundled R with
`<install_dir>/envs/mitopilot/bin/Rscript -e 'BiocManager::install("Smithsonian/MitoPilot@<version>")'`,
and in any other R you use for the app. When a release changes tool pins (see
`inst/native/VERSIONS.md` in the release), delete `<install_dir>/envs` and re-run the
install script.

## Troubleshooting

- **Solver errors during install**: run again with `--manager mamba` if you have
  one, or copy the failing `envs/*.yml` name into an issue.
- **`Could not find nextflow` in RStudio Server**: call `native_setup()`.
- **Jobs fail with `runmitos: command not found`**: the task did not source
  `activate.sh`; check the project `.config` has `process.beforeScript`.
- **Home quota exceeded**: choose an install location on a larger filesystem; the environments are large.
- **`'browser' must be a non-empty character string`**: an older MitoPilot in the
  bundled R; update it, or pass `launch.browser = FALSE`.
- **A different Java or Nextflow wins on PATH**: source `activate.sh` last,
  after any `module load`; it puts the bundled environment first.
- **`native_check()` shows the wrong Nextflow version**: look for an
  `NXF_VER` in `~/.Renviron` or your shell profile.
- **`sbatch: command not found` from the Start button in RStudio Server**: the
  install script did not see the scheduler when it ran (for example it ran on a
  node without one). Add `mp_path_add '/path/to/scheduler/bin'` to the job
  scheduler block in `activate.sh`, or run the install script again from a
  login node.

## Shared install by an administrator

The install script needs no root, only a writable directory, so an administrator can
run it once into shared space and every user sources the same `activate.sh`.
Three adjustments for a read-only install directory:

1. **Nextflow home must be writable per user.** `activate.sh` only sets
   `NXF_HOME` when it is unset, so users export their own (for example
   `~/.nextflow`) before sourcing, or the administrator edits that line.
   Nextflow writes plugins and engine downloads there.
2. **Permissions.** After the install, `chmod -R a+rX <install_dir>`. At run
   time everything writes into the project and the Nextflow work directory,
   never into the install directory. Prefer `micromamba`, `mamba`, or `conda`
   for a shared install: pixi's activation may try to repair an environment it
   considers stale, which fails on a read-only tree.
3. **The R package.** The administrator updates MitoPilot inside the bundled
   R; users may also install it in their own R for the app.

A module file is the natural wrapper. With Lmod:

```{bash eval=FALSE}
-- /apps/modulefiles/mitopilot/1.5.5.lua
setenv("MITOPILOT_NATIVE_PREFIX", "/apps/mitopilot_native")
setenv("NXF_HOME", pathJoin(os.getenv("HOME"), ".nextflow"))
execute { cmd = "source /apps/mitopilot_native/activate.sh", modeA = { "load" } }
```

Users then `module load mitopilot`, and `native_setup()` needs no argument.
`generate_config(container_engine = "none", native_prefix = "/apps/mitopilot_native")`
is unchanged for everyone.
