---
title: "Installing without containers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing without containers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
.alert {
  border-left: 5px solid;
  padding: 10px;
  margin: 10px 0;
  border-radius: 5px;
}
.alert-tip { border-color: #28A745; background-color: #E9F7EF; }
.alert-note { border-color: #007BFF; background-color: #EBF5FF; }
.alert-warning { border-color: #FFC107; background-color: #FFF9E6; }
.alert-danger { border-color: #DC3545; background-color: #F8D7DA; }
/* Smithsonian palette: blue #009CDE, yellow #FFCD00, black #101820 */
.alert-hydra {
  border-color: #101820;
  background-color: #E7F6FC;
  color: #101820;
}
.alert-hydra .hydra-tag {
  display: inline-block;
  background-color: #FFCD00;
  color: #101820;
  font-size: 0.75em;
  font-weight: bold;
  letter-spacing: 0.04em;
  padding: 1px 7px;
  border-radius: 10px;
  margin-right: 6px;
  vertical-align: 2px;
}
strong { font-weight: bold; }
img { max-width: 100%; border: 1px solid #ddd; border-radius: 4px; }
.figure { margin: 1.2em 0; }
.figure .caption, p.caption {
  font-size: 0.82em;
  line-height: 1.35;
  color: #6c757d;
  font-style: italic;
  margin-top: 0.4em;
}
.part-break {
  border: none;
  border-top: 3px solid #c9d1d6;
  margin: 2.2em 0 1.6em;
}
</style>

# Installing without containers

Some clusters do not allow Docker, Singularity, or Apptainer. MitoPilot can run
without them: every tool the pipeline needs installs into a folder you own with
a conda-style package manager, and no root access is required. This page sets
that up once per cluster.

<div class="alert alert-note">
<strong>What you get.</strong> The same pinned tool versions as the MitoPilot
container (R 4.5.2, Nextflow, GetOrganelle, MITOS2, BLAST+, and the rest), a
local copy of the metazoan BLAST database, and the MitoPilot R package, all under
one directory. Nothing uses user namespaces, ptrace, FUSE, or setuid binaries.
</div>

## Requirements

| Requirement | Notes |
|---|---|
| Linux x86_64, glibc 2.17 or newer | RHEL/Rocky/Alma 7+, Ubuntu 18.04+, Debian 10+ |
| About 20 GB of disk on a filesystem that allows executables | Home directories are often small or `noexec`; a project or scratch space is usually better |
| Outbound HTTPS from the node where you run the bootstrap | conda-forge, bioconda, GitHub, and NCBI are contacted once |
| `curl`, `tar`, `bash` | Present on every cluster |
| Optional: an existing `conda`, `mamba`, or `pixi` | If absent, the bootstrap downloads `micromamba` (a single binary) for you |

## 1. Run the bootstrap

Download the bootstrap script and the environment definitions from the MitoPilot
release you want (the script and the R package must be the same version):

```{bash eval=FALSE}
ver=1.5.5
curl -L https://github.com/Smithsonian/MitoPilot/archive/refs/tags/v${ver}.tar.gz | tar -xz
cd MitoPilot-${ver}/inst/native
```

Choose a prefix on a large, executable filesystem and run:

```{bash eval=FALSE}
bash bootstrap_native.sh --prefix /scratch/$USER/mitopilot_native
```

That builds the core environments (30 to 90 minutes, mostly downloads), fetches
the BLAST database, and installs the MitoPilot R package into the bundled R.
Options:

| Flag | Effect |
|---|---|
| `--manager pixi` | Use pixi instead of micromamba; `mamba` and `conda` also accepted |
| `--with-optional` | Also build MitoFinder, ARWEN, and NCBI ORFfinder (off by default in projects) |
| `--skip-blast-db` | Skip the 289 MB database; reference search falls back to NCBI over the network |
| `--mitopilot-ref v1.5.5` | Install a different MitoPilot tag |
| `--dry-run` | Show what would happen |

Re-running the script is safe; finished steps are skipped.

## 2. Check the environment

From any R that has MitoPilot installed (the bundled one is at
`<prefix>/envs/mitopilot/bin/R`):

```{r eval=FALSE}
library(MitoPilot)
native_check("/scratch/<user>/mitopilot_native")
```

Every core tool should show `found TRUE`. Optional tools are only present after
`--with-optional`.

## 3. Save a cluster config

```{r eval=FALSE}
generate_config(
  name             = "my_cluster",
  scheduler        = "slurm",             # or sge, pbs, lsf, local
  queue            = "general",
  account          = "my_allocation",
  container_engine = "none",
  native_prefix    = "/scratch/<user>/mitopilot_native"
)
```

Then create projects with `new_project(..., executor = "my_cluster")` exactly as
in [HPC cluster support](Custom-HPC.html). The config makes every pipeline task
source `<prefix>/activate.sh` before it runs, so no module loads or conda
activation are needed on compute nodes.

## 4. Launch the app

Which R runs the **app** does not matter, as long as MitoPilot is installed in it
and it is version 4.4 or newer. Pipeline tasks always use the bundled R.

**From a terminal (SSH tunnel).** Source the environment, start R, and follow
[Accessing the MitoPilot app over an SSH tunnel](Custom-HPC.html#accessing-the-mitopilot-app-over-an-ssh-tunnel):

```{bash eval=FALSE}
source /scratch/$USER/mitopilot_native/activate.sh
cd /scratch/$USER/my_runs/project01
R -e "MitoPilot::MitoPilot(host='0.0.0.0', port=7591, launch.browser=FALSE)"
```

**From RStudio Server or Open OnDemand.** Install MitoPilot in that R once
(`BiocManager::install("Smithsonian/MitoPilot")`), then in each session:

```{r eval=FALSE}
library(MitoPilot)
native_setup("/scratch/<user>/mitopilot_native")
```

`native_setup()` puts the bundled Nextflow and Java on this session's PATH. To
make it automatic, add `MITOPILOT_NATIVE_PREFIX=/scratch/<user>/mitopilot_native`
to `~/.Renviron` and call `native_setup()` with no arguments, or add the call to
`~/.Rprofile`. The **Start** button in the app reads the project config and finds
the environment on its own, so it works even without `native_setup()`; the setup
is what makes `library(MitoPilot)` report the right Nextflow version and lets you
run `nextflow` from the R console.

**Batch submission.** When the app offers a cluster submission script, the line
`source <prefix>/activate.sh` is already filled in.

## Updating

Update the R package inside the bundled R with
`<prefix>/envs/mitopilot/bin/Rscript -e 'BiocManager::install("Smithsonian/MitoPilot@v<version>")'`,
and in any other R you use for the app. When a release changes tool pins (see
`inst/native/VERSIONS.md` in the release), delete `<prefix>/envs` and re-run the
bootstrap.

## Troubleshooting

- **Solver errors during bootstrap**: run again with `--manager mamba` if you have
  one, or copy the failing `envs/*.yml` name into an issue.
- **`Could not find nextflow` in RStudio Server**: call `native_setup()`.
- **Jobs fail with `runmitos: command not found`**: the task did not source
  `activate.sh`; check the project `.config` has `process.beforeScript`.
- **Home quota exceeded**: set `--prefix` on scratch; the environments are large.
