---
title: "`r paste0('Submission Batch Review (', params$group, ')')`"
author: "MitoPilot"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    highlight: zenburn
params:
  db_path: ""
  group: ""
  start_aa: 10
  stop_aa: 10
  ident_pct: 60
---

```{r setup, include=FALSE, dev="CairoPNG"}
# force R to use the Cairo package to avoid issues rendering via R studio server
# see https://stackoverflow.com/questions/44731625/rmarkdown-with-no-x11
knitr::opts_chunk$set(echo = TRUE, dev = "CairoPNG")
```

```{r, echo=F, message=F, warning=F}
library(dplyr)
library(msaR)
```

```{r render_gene_aln, echo=F}
render_gene_aln <- function(gene) {
  gene_seqs <- annotations |>
    dplyr::filter(gene == !!gene) |>
    dplyr::pull(translation, seqid) |>
    Biostrings::AAStringSet()
  if (length(gene_seqs) < 2) {
    cat(paste0("_Gene **", gene, "** found in fewer than 2 samples, alignment skipped._\n"))
    return(invisible(NULL))
  }
  gene_aln <- DECIPHER::AlignSeqs(gene_seqs, processors = NULL, verbose = FALSE)
  dst <- DECIPHER::DistanceMatrix(gene_aln,
    includeTerminalGaps = TRUE, processors = NULL, type = "dist", verbose = FALSE
  )
  clust <- hclust(dst, "complete")
  msaR(
    gene_aln[clust$order],
    overviewbox = FALSE,
    seqlogo = FALSE,
    menu = FALSE,
    conservation = TRUE,
    labelNameLength = 150,
    colorscheme = "zappo",
    alignmentHeight = 400
  )
}
```

```{r, echo=F, message=T, warning=F}
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = params$db_path)
group <- params$group

# Collect PCG annotations with multi-exon genes merged (shared with the in-app
# outlier review; see MitoPilot:::get_export_PCG_annotations).
annotations <- MitoPilot:::get_export_PCG_annotations(con, group)
```

# Flagged Outliers

```{r flagged, echo=F, message=F, warning=F}
outliers <- MitoPilot::flag_PCG_outliers(
  group = group,
  db = params$db_path,
  start_aa = params$start_aa,
  stop_aa = params$stop_aa,
  ident_pct = params$ident_pct
)$flags

if (nrow(outliers) == 0) {
  cat("_No outlier PCG annotations were flagged._\n")
} else {
  outliers |>
    dplyr::transmute(
      Sample = label,
      Gene = toupper(gene),
      Issue = issue,
      `Start offset (aa)` = start_offset,
      `Stop offset (aa)` = stop_offset,
      `Identity (%)` = pct_identity
    ) |>
    knitr::kable()
}
```

Start/stop offset (aa) is the number of residues a sample's end extends past (+) or falls
short of (-) the alignment's well-aligned core. Flags use a start-offset threshold of
`r params$start_aa` aa, a stop-offset threshold of `r params$stop_aa` aa, and an identity
threshold of `r params$ident_pct`%.

# Internal Stop Codons

```{r internal_stops, echo=F, message=F, warning=F}
istops <- MitoPilot:::internal_stop_records(annotations)

if (nrow(istops) == 0) {
  cat("_No internal stop codons were found._\n")
} else {
  istops |>
    dplyr::transmute(
      Sample = label,
      Gene = toupper(gene),
      `Internal stops` = n_stops
    ) |>
    knitr::kable()
}
```

A protein-coding gene whose translation contains a stop codon before its end will
fail NCBI validation. Multi-exon genes are checked on the spliced translation.

# PCG Alignments

\* merged exons

*****

## NAD1

```{r nad1, echo=F, cache=F}
render_gene_aln("nad1")
```

## NAD2

```{r nad2, echo=F, cache=F}
render_gene_aln("nad2")
```

## COX1

```{r cox1, echo=F, cache=F}
render_gene_aln("cox1")
```

## COX2

```{r cox2, echo=F, cache=F}
render_gene_aln("cox2")
```

## ATP8

```{r atp8, echo=F, cache=F}
render_gene_aln("atp8")
```

## ATP6

```{r atp6, echo=F, cache=F}
render_gene_aln("atp6")
```

## COX3

```{r cox3, echo=F, cache=F}
render_gene_aln("cox3")
```

## NAD3

```{r nad3, echo=F, cache=F}
render_gene_aln("nad3")
```

## NAD4

```{r nad4, echo=F, cache=F}
render_gene_aln("nad4")
```

## NAD4l

```{r nad4l, echo=F, cache=F}
render_gene_aln("nad4l")
```

## NAD5

```{r nad5, echo=F, cache=F}
render_gene_aln("nad5")
```

## NAD6

```{r nad6, echo=F, cache=F}
render_gene_aln("nad6")
```

## CYTB

```{r cob, echo=F, cache=F}
render_gene_aln("cob")
```
